Reduced plasticity and relative variance in physiological rates of ectotherm populations under climate change

### 2024-06-13 ms.qmd
Updating responses to reviewer comments and changes MS. Make some notes on what to add after rendered to track changes version

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining phenotypic variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. We developed effect size estimates that quantify how physiological variance changes relative to the mean as temperature increases, which we refer to as 'relative variance'. We show by meta-analysing >1900 effects from 226 species that relative variance in physiological rates decreases at higher temperatures. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that beneficial acclimation responses in some ecosystems (marine and freshwater) may trade-off with reductions in physiological rate variance that can have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

We split the screening of titles and abstracts for the 1,321 papers found in our search among DWAN, FK, FS, and SN evenly. To ensure consistency among authors in title and abstract inclusion, relevant authors went through a randomly selected set of papers together before the formal screening to calibrate selection of papers based on our inclusion criteria (see below). In cases of disagreement regarding inclusion, we conservatively included the paper for full text screening and discussed uncertain papers among authors to come to a decision. After title and abstract screening, there was a total of 149 papers for full text screening. Papers were included only if they: 1) measured a physiological rate acutely at two temperatures on a sample of animals chronically exposed to the same two temperatures for at least 1 week; and 2) where physiological rates measured were burst and sustained locomotion, metabolic rates (standard, resting, routine and maximal), heart rates, and/or enzyme activities. We provide a PRISMA flow diagram of our extraction process in the *Supplement* (see @fig-prisma).

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
As indicated above, $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures. For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean or variance is larger at the higher of the two temperatures, standardized to 10&deg;C.

@@

@@

   model1.2_bayes <- brms::brm(effect_size | se(sqrt(V), sigma = TRUE) ~  max_temp_c + c_accl_time + type + habitat + type:habitat + (~1 | record_num) + (1 | gr(species_full, cov = A)) + (1 + max_temp_c| spp) + (1|trait), data2 = list(A=A), family = gaussian(), prior = c(prior(normal(0, 10), "Intercept"), prior(student_t(3, 0, 10), "sd")), control = list(adapt_delta = 0.98, max_treedepth = 12), chains = 4, thin = 5, iter = 5000, warmup = 1000, cores = 8, data = lnRRQ10_data)

@@

  model1.3_m_het <- metafor::rma.mv(effect_size ~  max_temp_c + c_accl_time + type + trait_category + type:trait_category, V = V, 
                              random = list(~1 | record_num,
                                            ~1 | species_full,
                                            ~1 | spp,
                                            ~1 | trait,
                                            ~1 + trait_category | obs),
                              R = list(species_full = m_phylo), 
                              struct = "HCS", rho = 0,  dfs = "contain", 
                              test = "t", data = lnRRQ10_data_m, control = list(iter.max=1000, rel.tol=1e-8))

@@

Relative variance in physiological rates ($lnCVR_{Q_{10}}$) showed a decrease with increasing temperature across all habitat types, which was especially pronounced in terrestrial and marine ectotherms. Overall, there was a `r mean(t_cvr_mean_ef)`% (95% CI: `r quantile(t_cvr_mean_ef, 0.025)[1]` to `r quantile(t_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(t_cvr_mean)`) reduction in relative physiological rate variance for terrestrial ectotherms and a `r mean(m_cvr_mean_ef)`% (95% CI: `r quantile(m_cvr_mean_ef, 0.025)[1]` to `r quantile(m_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(m_cvr_mean)`) reduction in relative variation for marine ectotherms when temperatures increased by 10$^\circ$C. In contrast, freshwater ectotherms exhibited a smaller reduction in relative physiological rate variance at high temperatures (`r mean(f_cvr_mean_ef)`%, 95% CI: `r quantile(f_cvr_mean_ef, 0.025)[1]` to `r quantile(f_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(f_cvr_mean)`).

@@

@@

$lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in relative variance) is log-linear. While $Q_{10}$ [@van1884etudes] has been used exstensively in the physiological literature to successfully address a multitude of questions [e.g., @Havird2020-mj; @Seebacher2015-wi], there is a preference for using a Boltzmann – Arrhenius (BA) relationship (or its more recent extension, the Sharpe-Schoolfield model [@molnar2017thermal; @michaletz2024hotter]) to model temperature effects on physiological rates [@gillooly2001effects; @michaletz2024hotter]. Over biologically realistic temperature ranges the BA and $Q_{10}$ approaches can vary at times by as much as 15% [@gillooly2001effects]. However, debate still exists over the utility of $Q_{10}$ when modelling temperature-dependence and both BA and $Q_{10}$ can exhibit curvilinearity at normal operating temperatures (as discussed in [@michaletz2024hotter]). @white2012information also showed that the BA model may not always perform better. For example, in eukaryotes the $Q_{10}$ approach provided a 5.8 fold better fit to metabolic rate data than the BA relationship [@white2012information].

Given that we restricted studies only to those that manipulated temperatures within normal operating ranges for a species and we only focus on eukaryotes, $Q_{10}$ should remain an appropriate way to capture temperature-dependent effects [@michaletz2024hotter]. However, given that it is not possible to collect full thermal performance curves for species within our data we could not validate this assumption directly. However, to correct mean effect size in our analysis as temperatures increase we included the maximum temperature of the experimental manipulations in our models. If the log-linearity assumption is violated then $lnRR_{Q_{10}}$ should decrease for studies applying higher temperature treatments because such temperatures are expected to either converge on or cross the thermal optima of the performance curve causing reaction rates to decelerate or decrease beyond $T_{op}$ [@michaletz2024hotter]. The benefit of such an approach means that we could still take advantage of the excellent properties of $Q_{10}$ as an effect size while statistically correcting for the potential non-linearity that would be expected in the data at high temperatures. Given that our data included a wide range of species and habitats, we also included a random slope of maximum temperature that varied across species because we, *a priori*, expected that each species will vary in their thermal performance curve and the chosen temperatures experimentally manipulated along the performance curve. To improve model interpretation we centred the maximum temperature around the mean temperature of the experimental treatments.

>>>>

Climate change is expected to result in warmer and more variable thermal environments globally. Maintaining phenotypic variability in physiological rates and adjusting them in response to extreme temperatures (plasticity) is essential for allowing populations to adapt to climate change. Yet, incorporating both plasticity and changes in phenotypic variation when predicting the impacts of climate change on populations is often ignored. We developed effect size estimates that quantify how physiological variance changes relative to the mean as temperature increases, which we refer to as 'relative variance'. We show by meta-analysing >1900 effects from 226 species that relative variance in physiological rates decreases at higher temperatures driven by the lower proportional increase in physiological rate variance relative to the mean. Freshwater ectotherms are capable of acclimating and have the smallest reductions in relative variance. Marine organisms also showed a capacity to acclimate to higher temperatures, but capacity for plasticity traded-off with a reduction in relative variance in physiological rates at higher temperatures. Relative variance reductions were particularly pronounced for terrestrial ectotherms, and this coincided with a lack of capacity for acclimation, highlighting the vulnerability of terrestrial ectotherms to climate change. Our results show that beneficial acclimation responses in some ecosystems (marine and freshwater) may trade-off with reductions in physiological rate variance that can have important evolutionary and ecological ramifications that affect our understanding of how climate change will impact populations now and in the future.

@@

We split the screening of titles and abstracts for the 1,321 papers found in our search among DWAN, FK, FS, and SN evenly. To ensure consistency among authors in title and abstract inclusion, relevant authors went through a randomly selected set of papers together before the formal screening to calibrate selection of papers based on our inclusion criteria (see below). In cases of disagreement regarding inclusion, we conservatively included the paper for full text screening and discussed uncertain papers among authors to come to a decision. After title and abstract screening, there was a total of 149 papers for full text screening. Papers were included only if they: 1) measured a physiological rate acutely at two temperatures on a sample of animals chronically exposed to the same two temperatures for at least 1 week; and 2) where physiological rates measured were burst and sustained locomotion, metabolic rates (standard, resting, routine and maximal), heart rates, and/or enzyme activities. Importantly, as done in @Seebacher2015-wi, we only included studies that manipulated temperatures within normal thermal ranges for the species. We provide a PRISMA flow diagram of our extraction process in the *Supplement* (see @fig-prisma).

@@

#### *Calculating acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*
As indicated above, $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ can all be taken from samples of organisms measured acutely at two temperatures or after having been acclimated these same temperatures. For studies that measure acute and acclimated responses we used the mean, standard deviation, and sample size to derive both acute and acclimation $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates. For studies that only measured $R_{1}$, $R_{2}$, $SD_{1}^2$ and $SD_{2}^2$ acutely we could only calculate acute versions of these effect size estimates. For all effect sizes the higher temperature was in the numerator and the lower of the two temperatures in the denominator. As such, positive effect sizes suggest that the mean or relative variance is larger at the higher of the two temperatures (numerator) when standardized to 10&deg;C.

#### *Assumptions of $lnRR_{Q_{10}}$, $lnVR_{Q_{10}}$ and $lnCVR_{Q_{10}}$ estimates*

$lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$, as with $Q_{10}$ more generally, all assume that the effect of temperature on physiological rates (or changes in relative variance) is log-linear. While this is likely in our data given that we restricted our analysis of $Q_{10}$ to standard operating temperatures for a given species, it may not always be satisfied given the diversity of species in our dataset. $Q_{10}$ [@van1884etudes] has been used exstensively in the physiological literature to successfully address a multitude of questions [e.g., @Havird2020-mj; @Seebacher2015-wi]. However, there is a preference for using a Boltzmann – Arrhenius (BA) relationship (or its more recent extension, the Sharpe-Schoolfield model [@molnar2017thermal; @michaletz2024hotter]) to model thermal effects on physiological rates [@gillooly2001effects; @michaletz2024hotter]. While debate still exists over the utility of $Q_{10}$ when modelling temperature-dependence it is important to recognise that both BA and $Q_{10}$ can exhibit curvilinearity as temperatures increase (as discussed in [@michaletz2024hotter]). @white2012information also showed that the BA model may not always perform better. For example, in eukaryotes, modelling thermal dependence using $Q_{10}$ provided a 5.8 fold better fit to metabolic rate data than the BA relationship [@white2012information]. Given that studies included in our analysis never measured full performance curves at acute and acclimation temperatures it was not possible for us to compare different models of thermal dependence. Nonetheless, $Q_{10}$-based effect sizes remain the most practical effect-size for comparing thermal dependence when using existing empircial data, with the benefit that these effects having convenient properties that make them suitable for meta-analysis. Nonetheless, we control for possible violations of the log-linearity assumption in our analyses (see below).

@@

In all our models we tested and corrected for the possibility of the log-linearity assumption of our effect sizes being violated. We predcited that, if $lnRR_{Q_{10}}$, $lnCVR_{Q_{10}}$ and $lnVR_{Q_{10}}$ were not strictly log-linear than we should see a decrease in average effect size for studies applying higher temperature treatments because these temperatures are expected to either converge on or cross the thermal optima of the performance curve causing reaction rates to decelerate or decrease beyond $T_{op}$ [@michaletz2024hotter]. The benefit of such an approach means that we could still take advantage of the excellent properties of $Q_{10}$ as an effect size while statistically correcting for the potential non-linearity that would be expected in the data at high temperatures. Given that our data included a wide range of species and habitats, we also included a random slope of maximum temperature that varied across species because we, *a priori*, expected that each species will vary in their thermal performance curve and the chosen temperatures experimentally manipulated along the performance curve. To improve model interpretation we centred the maximum temperature around the mean temperature of the experimental treatments.

@@

  # Secondary sensitivity analysis that uses cluster robust variance estimation to deal with unaccounted sources of non-independence. Not completely dealing with sharced sampling covariances in data as when collected these data were never indicated in past studies so hard to build full VCV matrix.

   model1.2_rve <- robust(model1.2, cluster = lnRRQ10_data$record_num) # Consistent.

  # Baysian model allows us to propagate error better is should be more robust to non-independence
   model1.2_bayes <- brms::brm(effect_size | se(sqrt(V), sigma = TRUE) ~  max_temp_c + c_accl_time + type + habitat + type:habitat + (~1 | record_num) + (1 | gr(species_full, cov = A)) + (1 + max_temp_c| spp) + (1|trait), data2 = list(A=A), family = gaussian(), prior = c(prior(normal(0, 10), "Intercept"), prior(student_t(3, 0, 10), "sd")), control = list(adapt_delta = 0.98, max_treedepth = 12), chains = 4, thin = 5, iter = 5000, warmup = 1000, cores = 8, data = lnRRQ10_data)

@@

  # Robust variance estimator.
    model1.3_m_rve <- robust(model1.3_m, cluster = lnRRQ10_data_m$record_num) # Mostly Consistent but some changes to p for some traits.

  # Baeysian model
  model1.3_m_het <- metafor::rma.mv(effect_size ~  max_temp_c + c_accl_time + type + trait_category + type:trait_category, V = V, 
                              random = list(~1 | record_num,
                                            ~1 | species_full,
                                            ~1 | spp,
                                            ~1 | trait,
                                            ~1 + trait_category | obs),
                              R = list(species_full = m_phylo), 
                              struct = "HCS", rho = 0,  dfs = "contain", 
                              test = "t", data = lnRRQ10_data_m, control = list(iter.max=1000, rel.tol=1e-8))

@@

Relative variance in physiological rates ($lnCVR_{Q_{10}}$) showed a decrease with increasing temperature across all habitat types, which was especially pronounced in terrestrial and marine ectotherms. Reductions in variance were driven by slower increases in the variance in physiological rates relative to the mean with scaling relationships between log standard deviation and log mean all being smaller than 1 and varying across habitats (See Supplementary Materials). Overall, there was a `r mean(t_cvr_mean_ef)`% (95% CI: `r quantile(t_cvr_mean_ef, 0.025)[1]` to `r quantile(t_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(t_cvr_mean)`) reduction in relative physiological rate variance for terrestrial ectotherms and a `r mean(m_cvr_mean_ef)`% (95% CI: `r quantile(m_cvr_mean_ef, 0.025)[1]` to `r quantile(m_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(m_cvr_mean)`) reduction in relative variation for marine ectotherms when temperatures increased by 10$^\circ$C. In contrast, freshwater ectotherms exhibited a smaller reduction in relative physiological rate variance at high temperatures (`r mean(f_cvr_mean_ef)`%, 95% CI: `r quantile(f_cvr_mean_ef, 0.025)[1]` to `r quantile(f_cvr_mean_ef, 0.975)[1]`, $p_{MCMC}$ = `r pmcmc(f_cvr_mean)`).

@@

#### Mean-variance relationships justify use of $lnCVR_{Q_{10}}$
We explored mean-variance relationships for the acute and acclimation responses across all traits and habitats. We estimated the scaling relationship between log standard deviation in physiological rates [log(SD)] and log mean in physiological rates [log(mean)], accounting for non-independence resulting from effects coming from the same species, study and traits (i.e., random effects of species, study and trait) as done in our main analyses. We also included an interaction between log(mean) and habitat type to better understand how the scaling relationship between log(SD) and log(mean) varies across habitats.

```{r, fig-mv}
#| label: fig-mv
#| fig-cap: "Mean-standard deviation relationships for the acute and acclimation responses across all habitats. Relationships are depicted for the low temperature treatment of the study (A) and high temperature treatment (B)."

# Model mean variance relationships to estimate the relationship
    model_mv1  <- brm(log(r1.1_sd) ~ log(r1.1)*habitat + (1|record_num) + (1|spp) + (1|trait), chains = 4, cores = 4, iter = 4000, warmup = 1000, thin = 1, control = list(adapt_delta = 0.98), data = lnRRQ10_data)
    model_mv2  <- brm(log(r1.2_sd) ~ log(r1.2)*habitat + (1|record_num) + (1|spp) + (1|trait), chains = 4, cores = 4, iter = 4000, warmup = 1000, thin = 1, control = list(adapt_delta = 0.98), data = lnRRQ10_data)

mv1 <- ggplot2::ggplot(lnRRQ10_data, aes(x = log(r1.1), y = log(r1.1_sd))) +
                geom_point(aes(color = habitat), size = 2.2) + 
                geom_smooth(method = "lm", se = FALSE) + 
                labs(x = "log (Mean)", y = "log (SD)", color = "Habitat") + 
                facet_wrap(~type) + theme_classic() + theme(axis.text = element_text(size=14), axis.title = element_text(size = 16), 
                legend.text = element_text(size = 14), legend.title = element_text(size = 16), strip.text = element_text(size = 16))

mv2 <- ggplot2::ggplot(lnRRQ10_data, aes(x = log(r1.2), y = log(r1.2_sd))) +
                geom_point(aes(color = habitat), size = 2.2) + 
                geom_smooth(method = "lm", se = FALSE) + 
                labs(x = "log (Mean)", y = "log (SD)", color = "Habitat") + 
                facet_wrap(~type) + theme_classic() + theme(axis.text = element_text(size=14), axis.title = element_text(size = 16), 
                legend.text = element_text(size = 14), legend.title = element_text(size = 16), strip.text = element_text(size = 16))

(mv1 / mv2)  + plot_annotation(tag_levels = "A", tag_suffix = ")") + theme(plot.tag = element_text(size = 20, face = "bold"))

Overall, we found that the relationship between the mean and SD of the acute and acclimation responses was generally linear, suggesting that the use of $lnCVR_{Q_{10}}$ was justified (@fig-mv). Overall, the scaling relationship between log(SD) and log(mean) was sub-linear across all habitats (Freshwater = $\Beta$ = `r fixef(model_mv1)[2]`, 95% CI: `r fixef(model_mv1)[2,3]` to  `r fixef(model_mv1)[2,4]`; Marine = $\Beta$ = `r fixef(model_mv1)[2] + fixef(model_mv1)[5]`, 95% CI: `r fixef(model_mv1)[2,3] + fixef(model_mv1)[5,3]` to `r fixef(model_mv1)[2,4] + fixef(model_mv1)[5,4]`; Terrestrial: $\Beta$ = `r fixef(model_mv1)[2] + fixef(model_mv1)[6]`, 95% CI: `r fixef(model_mv1)[2,3] + fixef(model_mv1)[6,3]` to `r fixef(model_mv1)[2,4] + fixef(model_mv1)[6,4]`) suggesting that the relative variance in physiological rates does not increase proportionally as the mean increases. Importantly, the scaling relationship between log(SD) and log(mean) differed significantly between habitats (Freshwater compared to Marine: Contrast = `r fixef(model_mv1)[5]`, 95% CI: `r fixef(model_mv1)[5,3]` to `r fixef(model_mv1)[5,4]` and Terrestrial to Marine: Contrast = `r fixef(model_mv1)[6]`, 95% CI: `r fixef(model_mv1)[6,3]` to `r fixef(model_mv1)[6,4]`).

@@
